---
title: "LRT"
author: "Mayleen"
date: "2026-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Org-level

First some pre-processing
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

eligible_counties <- readRDS("top1counties.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top) |> filter(geoid_2010 %in% eligible_counties)

df.changes <- df |> filter(topDamage) |> select(geoid_2010,TAX_YEAR)
tst <- as.list(df.changes$TAX_YEAR)
names(tst) <- df.changes$geoid_2010

df <- readRDS("../mega.rds") |> filter(TAX_YEAR <= 2019) # 13,211,506 records; 1,235,390 organizations

# Adjust dollar amounts for inflation to the year 2019
adj2019 <- as.data.frame(read_csv("../usd_2019_conv.csv", show_col_types = FALSE))
df <- merge(df, adj2019, by.x = "TAX_YEAR", by.y = "YEAR")
df <- df |> mutate(TOT_ASSET_OG = TOT_ASSET,
                   TOT_REV_OG = TOT_REV,
                   TOT_EXP_OG = TOT_EXP,
                   TOT_ASSET = TOT_ASSET * ADJ_2019,
                   TOT_REV = TOT_REV * ADJ_2019,
                   TOT_EXP = TOT_EXP * ADJ_2019)
# Optional, just to keep minimal number of columns
df <- df |> select(-TOT_ASSET_OG, -TOT_REV_OG, -TOT_EXP_OG, -DATA_COUNT)

# July 14, 2026: if negative, convert to 0
df <- df |> mutate(TOT_REV = ifelse(TOT_REV < 0, 0, TOT_REV),
                   TOT_ASSET = ifelse(TOT_ASSET < 0, 0, TOT_ASSET),
                   TOT_EXP = ifelse(TOT_EXP < 0, 0, TOT_EXP))

# IDs for two counties were changed after 2010. Some datasets use latest ID, others use 2010 ID
df <- df |> mutate(geoid_2010 = county.census.geoid)
df$geoid_2010[df$county.census.geoid == "02158"] <- "02270"
df$geoid_2010[df$county.census.geoid == "46102"] <- "46113"

# Now we need to figure out which orgs we actually will be able to use
# For sure if they have less than 10 records they won't be eligible
df <- df |> filter(geoid_2010 %in% eligible_counties) |> filter(!is.na(TOT_REV)) # 496,587 records from 48,300 orgs
df <- df |> group_by(EIN2) |> 
      mutate(DATA_COUNT = n()) |> 
      ungroup() |> filter(DATA_COUNT >= 10) # 401,352 records from 21,296 orgs

# Now need to make sure there is an appropriate number of records before and after the suspected change point
# First, let's make sure that we are only looking at orgs that have an existing record during the candidate change point year
df <- merge(df, df.changes |> mutate(topDamage = TRUE), 
            all.x = TRUE, by = c("geoid_2010", "TAX_YEAR"))
df <- df |> mutate(topDamage = replace_na(topDamage, FALSE))

eligible_orgs <- df |> group_by(EIN2) |> 
      summarize(eligible = (sum(topDamage)) > 0) |> 
      filter(eligible) |>
      pull(EIN2) # 15,573 orgs 

df <- df |> filter(EIN2 %in% eligible_orgs) # 316,854 records from 15,573 orgs 

# side quest resolving duplicates.. just removing all records in question for now
bad_groups <- df |> group_by(EIN2, TAX_YEAR) |>
      summarize(n = n(), .groups = "keep") |>
      filter(n > 1) # 562 org,year pairs that are unresolved dupes
bad_groups <- bad_groups |>  select(-n) |> mutate(keep = FALSE)

df <- merge(df, bad_groups, all.x = TRUE, by = c("EIN2", "TAX_YEAR")) |>
      mutate(keep = replace_na(keep, TRUE)) |>
      filter(keep) # 315,729 records from 15,573 orgs 

# Now, let's make sure there is an appropriate buffer around the change point
gaps <- df |> 
      group_by(EIN2) |>
      arrange(TAX_YEAR) |>
      summarise(before = sum(row_number() < which(topDamage)),
             after = sum(row_number() > which(topDamage)))

eligible_orgs <- gaps |> filter(before >= 5 & after >= 4) |> pull(EIN2) # 11,164 orgs

saveRDS(df |> filter(EIN2 %in% eligible_orgs), "df_eligible_orgs.rds") # 246,231 records
```

## Hyperparameter Optimization
### Grid Search
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

set.seed(2727)
orgs_sample <- sample(unique(df$EIN2), size = 50)

combos <- expand.grid(nu = seq(from = 0.01, to = 3, by = 0.07475), 
                      nugget = seq(from = 0, to = 3, by = 0.1))

best_parameters.null <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))

for (orgID in orgs_sample) {
      # pull their time series
      my_data <- df |> filter(EIN2 == orgID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      # grid search (if doing Nelder-Mead, multiply get_likelihood_sigma by -1)
      likelihood_wrapper.null <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      likelihoods.null <- apply(combos, MARGIN = 1, likelihood_wrapper.null)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.null))
      best_parameters.null <- rbind(best_parameters.null, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      likelihoods.L <- apply(combos, MARGIN = 1, likelihood_wrapper.left)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.L))
      best_parameters.L <- rbind(best_parameters.L, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      likelihoods.R <- apply(combos, MARGIN = 1, likelihood_wrapper.right)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.R))
      best_parameters.R <- rbind(best_parameters.R, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
}

all_best <- rbind(best_parameters.null, best_parameters.L, best_parameters.R) |> select(nu,nugget)
cl <- kmeans(all_best, 5, iter.max = 20, nstart = 3)
# saveRDS(cl, "ORG_init_conds_50.rds")

cl$centers
```

### Nelder-Mead
Note: This took a little over 4 hours to run
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

init_conds <- readRDS("ORG_init_conds_50.rds")$centers

best_parameters.null <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ein", "nu", "nugget"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ein", "nu", "nugget"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ein", "nu", "nugget"))

errs <- c()
for (orgID in unique(df$EIN2)) {
      print(orgID)
      # pull their time series
      my_data <- df |> filter(EIN2 == orgID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      likelihood_wrapper.null <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
i <- 2      
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.null,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }

      best_parameters.null <- rbind(best_parameters.null, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      if (var(centered_rev.L) == 0){
            errs <- c(errs, orgID)
            next
            }
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.left,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      best_parameters.L <- rbind(best_parameters.L, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      if (var(centered_rev.R) == 0){
            errs <- c(errs, orgID)
            next
            }
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.right,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      best_parameters.R <- rbind(best_parameters.R, 
                                    as.data.frame(list(ein = orgID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))

}

saveRDS(best_parameters.null, "ORG_best_null.rds")
saveRDS(best_parameters.L, "ORG_best_L.rds")
saveRDS(best_parameters.R, "ORG_best_R.rds")
```

## LRT
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

best_all.null <- readRDS("ORG_best_null.rds")
best_all.L <- readRDS("ORG_best_L.rds")
best_all.R <- readRDS("ORG_best_R.rds")
missing_ein <- union(setdiff(unique(df$EIN2), best_all.L$ein),
                     setdiff(unique(df$EIN2), best_all.R$ein))

# saveRDS(missing_ein, "org_missing_ein.rds")

rho = 1

reject_null <- list()
for (my_org in unique(df$EIN2)) {
      if (my_org %in% missing_ein){next}
      # pull their time series
      my_data <- df |> filter(EIN2 == my_org) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      
      # pull best parameters
      best.null <- best_all.null |> filter(ein == my_org)
      best.L <- best_all.L |> filter(ein == my_org)
      best.R <- best_all.R |> filter(ein == my_org)
      
      
      # Determine the left and right time series corresponding to the current candidate change point
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Hypothesis Model ###
      curr_all_years <- c(left_years, right_years) # Redundant for single candidate
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years]) # center the log revenue values so they are mean 0
      
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      Sigma_0 <- Matern(d = dist_mat,
                        smoothness = best.null$nu,
                        range = rho,
                        phi = 1) + diag(best.null$nugget, dim(dist_mat)[1])
      
      sigma.sqrd_0 <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(Sigma_0)) %*% centered_rev)[1,1]
      Sigma_0 <- sigma.sqrd_0 * Sigma_0
      
      ### Alternative Hypothesis Model ###
      # center the log revenue values so they are mean 0
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      
      # compute and normalize distance matrices
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      # Compute Matern covariance matrices
      Sigma_L <- Matern(d = dist_mat.L,
                        smoothness = best.L$nu,
                        range = rho,
                        phi = 1) + diag(best.L$nugget, dim(dist_mat.L)[1])
      sigma.sqrd_L <- (1/length(left_years)) * (centered_rev.L %*% chol2inv(chol(Sigma_L)) %*% centered_rev.L)[1,1]
      Sigma_L <- sigma.sqrd_L * Sigma_L
      
      Sigma_R <- Matern(d = dist_mat.R,
                        smoothness = best.R$nu,
                        range = rho,
                        phi = 1) + diag(best.R$nugget, dim(dist_mat.R)[1])
      sigma.sqrd_R <- (1/length(right_years)) * (centered_rev.R %*% chol2inv(chol(Sigma_R)) %*% centered_rev.R)[1,1]
      Sigma_R <- sigma.sqrd_R * Sigma_R
      
      ### Likelihood Ratio Test ###
      L_0 <- dmvnorm(centered_rev, 
                     mean = rep(0, length(centered_rev)), 
                     sigma = Sigma_0,
                     log = TRUE)
      
      L_left <- dmvnorm(centered_rev.L, 
                        mean = rep(0, length(centered_rev.L)), 
                        sigma = Sigma_L,
                        log = TRUE)
      
      L_right <- dmvnorm(centered_rev.R, 
                         mean = rep(0, length(centered_rev.R)), 
                         sigma = Sigma_R,
                         log = TRUE)
      
      L_alt <- L_left + L_right
      
      
      # likelihood ratio test statistic
      lrt <- -2*L_0 + 2*L_alt
      doF <- 4 #  (mu1, mu2, Sigma1, Sigma2, nu1, nu2, nugget1, nugget2) vs (mu, Sigma, nu, nugget) 
      reject_null[my_org] <- (lrt > qchisq(p = 0.05, df = doF, lower.tail = FALSE))
}

sum(unlist(reject_null)) / length(reject_null) # 37.25 % of the time we have enough evidence to reject the null
```



# Empirical Distribution Org-Level
```{r}
library(readr)
library(dplyr)
library(tidyverse)

set.seed(2727)
df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

df <- df |> filter(!(geoid_2010 %in% readRDS("org_missing_ein.rds"))) |> 
      group_by(EIN2) |>
      mutate(n = n(),
             row_num = row_number(),
             eligible = (row_num > 5 & row_num <= n-4)) |> ungroup()

candidates_all <- list()
for (i in 1:50) {
      candidates <- df |> group_by(EIN2) |>
            filter(eligible) |>
            slice_sample(n = 1) |>
            select(EIN2, TAX_YEAR)
      
      candidates_list <- candidates |> pull(TAX_YEAR) |> as.list()
      names(candidates_list) <- candidates |> pull(EIN2)
      
      candidates_all[[as.character(i)]] <- candidates_list
}
# saveRDS(candidates_all, "ORG_random_candidates50.rds")
```

## Hyperparameter Optimization
Grid Search
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)
candidates_all <- readRDS("ORG_random_candidates50.rds")

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

set.seed(2727)
orgs_sample <- sample(unique(df$EIN2), size = 50)

combos <- expand.grid(nu = seq(from = 0.01, to = 3, by = 0.07475), 
                      nugget = seq(from = 0, to = 3, by = 0.1))

best_parameters.null <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ein", "nu", "nugget", "trial"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ein", "nu", "nugget", "trial"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ein", "nu", "nugget", "trial"))

for (orgID in orgs_sample) {
      # pull their time series
      my_data <- df |> filter(EIN2 == orgID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      for (i in names(candidates_all)){      
            candidate <- candidates_all[[i]][[orgID]] - min(df$TAX_YEAR) + 1
            left_years <- all_years[1:which(all_years == candidate)-1]
            right_years <- all_years[which(all_years == candidate):length(all_years)]
            
            ### Null Model ###
            curr_all_years <- c(left_years, right_years)
            centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
            dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
            dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
            
            # grid search (if doing Nelder-Mead, multiply get_likelihood_sigma by -1)
            likelihood_wrapper.null <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat,
                                                   rho = 1,
                                                   data = centered_rev))
            }
            
            likelihoods.null <- apply(combos, MARGIN = 1, likelihood_wrapper.null)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.null))
            best_parameters.null <- rbind(best_parameters.null, 
                                          as.data.frame(list(ein = orgID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
            
            ### Alternative Model ###
            
            # Left
            centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
            dist_mat.L <- dist_mat_all[left_years, left_years]
            dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
            
            likelihood_wrapper.left <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat.L,
                                                   rho = 1,
                                                   data = centered_rev.L))
            }
            
            likelihoods.L <- apply(combos, MARGIN = 1, likelihood_wrapper.left)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.L))
            best_parameters.L <- rbind(best_parameters.L, 
                                          as.data.frame(list(ein = orgID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
            
            # Right
            centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
            dist_mat.R <- dist_mat_all[right_years, right_years]
            dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
            
            likelihood_wrapper.right <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat.R,
                                                   rho = 1,
                                                   data = centered_rev.R))
            }
            
            likelihoods.R <- apply(combos, MARGIN = 1, likelihood_wrapper.right)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.R))
            best_parameters.R <- rbind(best_parameters.R, 
                                          as.data.frame(list(ein = orgID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
            }
}

all_best <- rbind(best_parameters.null, best_parameters.L, best_parameters.R) |> select(nu,nugget)
cl <- kmeans(all_best, 5, iter.max = 20, nstart = 3)
# saveRDS(cl, "ORG_empirical_init_conds_50.rds")

cl$centers
```

Nelder-Mead
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
library(doParallel)
source("helper.R")

candidates_all <- readRDS("ORG_random_candidates50.rds")
df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

init_conds <- readRDS("ORG_empirical_init_conds_50.rds")$centers

best_parameters <- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ein", "nu", "nugget", "trial"))

errs <- c()
combos <- expand.grid(unique(df$EIN2), as.character(1:length(candidates_all))) |> arrange(Var1, Var2)

test_idx <- c(1:10, 51:60)
resBest <- foreach(i = test_idx, .combine = 'rbind', .packages = c("dplyr", "fields", "mvtnorm"), .verbose = TRUE, .errorhandling = "remove") %do% {
      orgID <- combos$Var1[i]
      trialID <- combos$Var2[i]
      print(paste("EIN:", orgID, "Trial Number:", trialID))
      # pull their time series
      my_data <- df |> filter(EIN2 == orgID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- candidates_all[[trialID]][[orgID]] - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      likelihood_wrapper.null <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
i <- 2      
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.null,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      res.null <- as.data.frame(list(type = "null",
                         ein = orgID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      if (var(centered_rev.L) == 0){
            cat(i, file = file.path("errors/", paste0(i, ".txt")))
            }
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.left,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      res.left <- as.data.frame(list(type = "left",
                         ein = orgID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      if (var(centered_rev.R) == 0){
            cat(i, file = file.path("errors/", paste0(i, ".txt")))
            }
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.right,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      rbind(res.null,
            res.left,
            as.data.frame(list(type = "right",
                         ein = orgID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID)))

}

# saveRDS(resBest, "empirical_best_test.rds")

```

## LRT
```{r}
library(readr)
library(dplyr)
library(tidyverse)

glued <- rbind(readRDS("ORG_empirical_best_1.rds"), 
               readRDS("ORG_empirical_best_2.rds"),
               readRDS("ORG_empirical_best_3.rds"),
               readRDS("ORG_empirical_best_4.rds"),
               readRDS("ORG_empirical_best_5.rds"))

# Creating a list of lists
best_all <- glued |> 
      mutate(ID = paste0(ein,".",trial)) |> 
      select(-ein,-trial) # First, creating an ID for each org,trial pair
best_all <- split(best_all, best_all$ID) # The outer list names are these IDs
best_all <- lapply(best_all, function(x) {split(x,x$type)}) # The inner list names are null, left, right with elements being the values nu and nugget
# Example: best_all$ID$left$nu

saveRDS(best_all, "ORG_best_empirical_all.rds")
```


```{r}
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
library(doParallel)

candidates_all <- readRDS("ORG_random_candidates50.rds")
df <- readRDS("df_eligible_orgs.rds") |> select(-DATA_COUNT, -keep)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

best_all <- readRDS("ORG_best_empirical_all.rds")

rho = 1

# note: .inorder = FALSE so the outer list won't be in trial order... I think for our purposes ok
res <- foreach(trialID = 1:3, .packages = c("dplyr", "fields", "mvtnorm"), .inorder = FALSE) %dopar% { # length(candidates_all)
      reject_null <- list()
      print(paste0("Trial: ", trialID))
      for (orgID in unique(df$EIN2)){
            ID <- paste0(orgID, ".", trialID)
            if (!(ID %in% names(best_all))){
                  print(paste("Missing ID:", ID))
                  next
            }
            # pull their time series
            my_data <- df |> filter(EIN2 == orgID) |>
                  mutate(TOT_REV = log1p(TOT_REV),
                         TOT_EXP = log1p(TOT_EXP),
                         TOT_ASSET = log1p(TOT_ASSET))
            all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
            log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
            
            candidate <- candidates_all[[trialID]][[orgID]] - min(df$TAX_YEAR) + 1
            
            # pull best parameters
            best.curr <- best_all[[ID]] # best.curr$left$nu
      
            # Determine the left and right time series corresponding to the current candidate change point
            left_years <- all_years[1:which(all_years == candidate)-1]
            right_years <- all_years[which(all_years == candidate):length(all_years)]
            
            ### Null Hypothesis Model ###
            curr_all_years <- c(left_years, right_years) # Redundant for single candidate
            centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years]) # center the log revenue values so they are mean 0
            
            dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
            dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
            Sigma_0 <- Matern(d = dist_mat,
                              smoothness = best.curr$null$nu,
                              range = rho,
                              phi = 1) + diag(best.curr$null$nugget, dim(dist_mat)[1])
            
            sigma.sqrd_0 <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(Sigma_0)) %*% centered_rev)[1,1]
            Sigma_0 <- sigma.sqrd_0 * Sigma_0
            
            ### Alternative Hypothesis Model ###
            # center the log revenue values so they are mean 0
            centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
            centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
            
            # compute and normalize distance matrices
            dist_mat.L <- dist_mat_all[left_years, left_years]
            dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
            
            dist_mat.R <- dist_mat_all[right_years, right_years]
            dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
            
            # Compute Matern covariance matrices
            Sigma_L <- Matern(d = dist_mat.L,
                        smoothness = best.curr$left$nu,
                        range = rho,
                        phi = 1) + diag(best.curr$left$nugget, dim(dist_mat.L)[1])
            sigma.sqrd_L <- (1/length(left_years)) * (centered_rev.L %*% chol2inv(chol(Sigma_L)) %*% centered_rev.L)[1,1]
            Sigma_L <- sigma.sqrd_L * Sigma_L
            
            Sigma_R <- Matern(d = dist_mat.R,
                        smoothness = best.curr$right$nu,
                        range = rho,
                        phi = 1) + diag(best.curr$right$nugget, dim(dist_mat.R)[1])
            sigma.sqrd_R <- (1/length(right_years)) * (centered_rev.R %*% chol2inv(chol(Sigma_R)) %*% centered_rev.R)[1,1]
            Sigma_R <- sigma.sqrd_R * Sigma_R
            
            ### Likelihood Ratio Test ###
            L_0 <- dmvnorm(centered_rev, 
                           mean = rep(0, length(centered_rev)), 
                           sigma = Sigma_0,
                           log = TRUE)
            
            L_left <- dmvnorm(centered_rev.L, 
                              mean = rep(0, length(centered_rev.L)), 
                              sigma = Sigma_L,
                              log = TRUE)
            
            L_right <- dmvnorm(centered_rev.R, 
                               mean = rep(0, length(centered_rev.R)), 
                               sigma = Sigma_R,
                               log = TRUE)
            
            L_alt <- L_left + L_right
      
            
            # likelihood ratio test statistic
            lrt <- -2*L_0 + 2*L_alt
            doF <- 4 #  (mu1, mu2, Sigma1, Sigma2, nu1, nu2, nugget1, nugget2) vs (mu, Sigma, nu, nugget)
            reject_null[[orgID]] <- (lrt > qchisq(p = 0.05, df = doF, lower.tail = FALSE))

      }
      # reject_null_all[[as.character(trialID)]] <- reject_null
      reject_null
}

res.final <- unlist(lapply(res, function(x){sum(unlist(x)) / length(x)}))
hist(res.final)
sd(res.final)
```




## Bootstrap
```{r}
library(readr)
library(dplyr)
library(tidyverse)
library(doParallel)

res <- readRDS("org_empirical_lrt_res.rds")

# testing that this works
# tst <- unlist(res)
# tst <- split(tst, names(tst))
# tst2 <- unname(tst$`EIN-01-0245341`)
# tst3 <- c()
# for (i in 1:length(res)){
#       tst3 <- c(tst3, res[[i]]$`EIN-01-0245341`)
# }
# all(tst2 == tst3)
# tst <- lapply(tst, unname)

res <- unlist(res)
res <- split(res, names(res))
res <- lapply (res, unname)

empirical <- foreach(i = 1:1000, .inorder = FALSE) %do% {
      tst <- lapply(res, function(x){sample(x, 1)})
      sum(unlist(tst)) / length(tst)
}

hist(unlist(empirical))
quantile(unlist(empirical), probs = c(0.025, 0.5, 0.975)) # middle 95% of data is between 37.27% and 38.45%
mean(unlist(empirical))
```

# County-level
First, some data exploration and setup. 
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)

df <- readRDS("../county_long.rds") |> filter(TAX_YEAR < 2020)
top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile

cols_to_keep <- c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")

df <- df |> select(all_of(cols_to_keep))

df <- df |> mutate(topDamage = TotPerCapADJ >= top)
sum(df$topDamage) # 905

# counties are only eligible if at least one significant damage event occurred in their time series
eligible_counties <- df |> filter(topDamage) |> select(geoid_2010) |> unique() |> pull() # 657 eligible counties
df <- df |> filter(geoid_2010 %in% eligible_counties)

# Determine if significant damage events (i.e. being in top 1 percentile) occur in consecutive years 
df <- df |> group_by(geoid_2010) |> 
      arrange(TAX_YEAR, .by_group = TRUE) |>
      mutate(prev = lag(topDamage)) |>
      ungroup()

top_summary <- df |>
      group_by(geoid_2010) |>
      filter(topDamage) |>
      summarise(n_top = sum(topDamage),
                n_consecutive = sum(prev))

table(top_summary$n_top) # 492 counties with 1 change point, 113 with 2, 52 with 3 or more
table(top_summary$n_consecutive) # 615 with no consecutive, 37 with 1 or 2 consecutive
sum(is.na(df |> filter(topDamage) |> select(prev))) # 5; all due to being at beginning of time series
df |> filter(topDamage & is.na(prev) & TAX_YEAR != 1991) # 1 county whose time series began after 1991

####### 
# Old code looking at gaps and choosing candidates based on there being enough buffer around
# Determine gaps between significant damage events
# gaps <- df |> group_by(geoid_2010) |>
#       select(TAX_YEAR, geoid_2010, top10damage) |>
#       mutate(min_year = min(TAX_YEAR),
#              max_year = max(TAX_YEAR)) |>
#       filter(top10damage) |>
#       mutate(gap_before = if_else(is.na(TAX_YEAR-lag(TAX_YEAR)), TAX_YEAR-min_year, TAX_YEAR-lag(TAX_YEAR)-1),
#              gap_after = if_else(is.na(lead(TAX_YEAR)-TAX_YEAR), max_year-TAX_YEAR, lead(TAX_YEAR)-TAX_YEAR-1))
# 
# sum(gaps$gap_before >= 5 & gaps$gap_after >= 5) # 1580 
# 
# sum(gaps$gap_before >= 4 & gaps$gap_after >= 4) # 2203
# 
# sum(gaps$gap_before >= 5 & gaps$gap_after >= 4) # 1826 -- since including change point in second set, still get 5 points "after"
# 
# eligible_counties <- gaps |> filter(gap_before >= 5 & gap_after >= 4) |> select(geoid_2010) |> unique() |> pull() # 1507
#
# top_summary |> filter(geoid_2010 %in% eligible_counties) |> select(n_top) |> pull() |> table()
# 
# top_summary |> filter(geoid_2010 %in% eligible_counties) |> select(n_consecutive) |> pull() |> table() 
# 
# df <- df |> filter(geoid_2010 %in% eligible_counties) |>
#       merge(gaps |> mutate(candidate_change_pt = (gap_before >= 5 & gap_after >= 4)) |> 
#       select(TAX_YEAR, geoid_2010, candidate_change_pt),
#       by = c("TAX_YEAR", "geoid_2010"), all.x = TRUE) |>
#       arrange(geoid_2010,TAX_YEAR)
# 
# df <- df |> mutate(candidate_change_pt = replace_na(candidate_change_pt,FALSE))
# 
# df |> group_by(geoid_2010) |> summarize(n_change_pts = sum(candidate_change_pt)) |> select(n_change_pts) |> pull() |> table() # 1209 with 1, 278 with 2, 19 with 3, 1 with 4
#######

# keep only the time series with one candidate change point
eligible_counties <- top_summary |> filter(n_top == 1) |> select(geoid_2010) |> pull() # 492
df <- df |> filter(geoid_2010 %in% eligible_counties)

gaps <- df |> group_by(geoid_2010) |>
      select(TAX_YEAR, geoid_2010, topDamage) |>
      mutate(min_year = min(TAX_YEAR),
             max_year = max(TAX_YEAR)) |>
      filter(topDamage) |>
      mutate(gap_before = TAX_YEAR-min_year, gap_after = max_year-TAX_YEAR)

# Keep only candidates that are not too close to the beginning or end of their time series
# TODO: to really catch the right buffer length I also need to check that there actually exist 5 data points before and 4 after; currently we check the distance between the minimum and maximum year but if there is missing data in between then it would not be equivalent
eligible_counties <- gaps |> filter(gap_before >= 5 & gap_after >= 4) |> pull(geoid_2010)

df <- df |> filter(geoid_2010 %in% eligible_counties)
# saveRDS(eligible_counties, "top1counties.rds")
```

## Multiple Candidates
```{r}
for (idx in seq(length(candidate))) { 
      # Determine the left and right time series corresponding to the current candidate change point
      if (length(candidate) == 1){ # only a single candidate point
            left_years <- all_years[1:which(all_years == candidate)-1]
            right_years <- all_years[which(all_years == candidate):length(all_years)]
      } else if (idx == 1){ # more than one candidate point, this is the first
            # TODO: for this & remaining cases need to replace candidate[idx] with which(all_years == candidate[idx])
            # left data points
            left_years <- all_years[1:candidate[1]-1]
            right_years <- all_years[candidate[1]:(candidate[2]-1)]
      } else if (idx == length(candidate)) { # more than one candidate point, this is the last
            left_years <- all_years[candidate[idx-1]:(candidate[idx]-1)]
            right_years <- all_years[candidate[idx]:length(all_years)]
      } else { # more than one candidate point, this is the first, this is in the middle
            left_years <- all_years[candidate[idx-1]:(candidate[idx]-1)]
            right_years <- all_years[candidate[idx]:(candidate[idx+1]-1)]
      }
}
```

## Hyperparameter Optimization

### Single county, single change point
```{r}
library(fields)
library(mvtnorm)
source("helper.R")

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

countyID <- "01061"

# pull their time series
my_data <- df |> filter(geoid_2010 == countyID) |>
      mutate(TOT_REV = log1p(TOT_REV),
             TOT_EXP = log1p(TOT_EXP),
             TOT_ASSET = log1p(TOT_ASSET))
min_year <- min(my_data$TAX_YEAR)
all_years <- my_data$TAX_YEAR - min_year + 1 # all the years we have data for this county
log_rev <- my_data$TOT_REV

candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min_year + 1
left_years <- all_years[1:candidate-1]
right_years <- all_years[candidate:length(all_years)]
combos <- expand.grid(nu = seq(from = 0.01, to = 3, by = 0.1196), 
                      nugget = seq(from = 0, to = 3, by = 0.15))

### Null Model ###
curr_all_years <- c(left_years, right_years)
centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))

# grid search (if doing Nelder-Mead, multiply get_likelihood_sigma by -1)
likelihood_wrapper.null <- function(pars.vec){
      return(get_likelihood_sigma(row = pars.vec,
                                       dist_mat = dist_mat,
                                       rho = 1,
                                       data = centered_rev))
}

likelihoods.null <- apply(combos, MARGIN = 1, likelihood_wrapper.null)
tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.null))

best_parameters.null <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.null <- rbind(best_parameters.null, 
                              as.data.frame(list(geoid = countyID,
                                                 nu = tst[which.max(tst$likelihood),]$nu,
                                                 nugget = tst[which.max(tst$likelihood),]$nugget)))

### Alternative Model ###

# Left
centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
dist_mat.L <- dist_mat_all[left_years, left_years]
dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))

likelihood_wrapper.left <- function(pars.vec){
      return(get_likelihood_sigma(row = pars.vec,
                                       dist_mat = dist_mat.L,
                                       rho = 1,
                                       data = centered_rev.L))
}

likelihoods.L <- apply(combos, MARGIN = 1, likelihood_wrapper.left)
tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.L))

best_parameters.L <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.L <- rbind(best_parameters.L, 
                              as.data.frame(list(geoid = countyID,
                                                 nu = tst[which.max(tst$likelihood),]$nu,
                                                 nugget = tst[which.max(tst$likelihood),]$nugget)))

# Right
centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
dist_mat.R <- dist_mat_all[right_years, right_years]
dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))

likelihood_wrapper.right <- function(pars.vec){
      return(get_likelihood_sigma(row = pars.vec,
                                       dist_mat = dist_mat.R,
                                       rho = 1,
                                       data = centered_rev.R))
}

likelihoods.R <- apply(combos, MARGIN = 1, likelihood_wrapper.right)
tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.R))

best_parameters.R <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.R <- rbind(best_parameters.R, 
                              as.data.frame(list(geoid = countyID,
                                                 nu = tst[which.max(tst$likelihood),]$nu,
                                                 nugget = tst[which.max(tst$likelihood),]$nugget)))
saveRDS(best_parameters.null, "best_null.rds")
saveRDS(best_parameters.L, "best_L.rds")
saveRDS(best_parameters.R, "best_R.rds")
```

### Multiple counties, single change point - Grid Search w/ 50 counties
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

df <- readRDS("../county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

eligible_counties <- readRDS("top1counties.rds")
eligible_counties <- sample(eligible_counties, size = 50)

combos <- expand.grid(nu = seq(from = 0.01, to = 3, by = 0.07475), 
                      nugget = seq(from = 0, to = 3, by = 0.1))

best_parameters.null <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
for (countyID in eligible_counties) {
      # pull their time series
      my_data <- df |> filter(geoid_2010 == countyID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      # grid search (if doing Nelder-Mead, multiply get_likelihood_sigma by -1)
      likelihood_wrapper.null <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      likelihoods.null <- apply(combos, MARGIN = 1, likelihood_wrapper.null)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.null))
      best_parameters.null <- rbind(best_parameters.null, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      likelihoods.L <- apply(combos, MARGIN = 1, likelihood_wrapper.left)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.L))
      best_parameters.L <- rbind(best_parameters.L, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      likelihoods.R <- apply(combos, MARGIN = 1, likelihood_wrapper.right)
      tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.R))
      best_parameters.R <- rbind(best_parameters.R, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
}

all_best <- rbind(best_parameters.null, best_parameters.L, best_parameters.R) |> select(nu,nugget)
cl <- kmeans(all_best, 5, iter.max = 20, nstart = 3)
saveRDS(cl, "init_conds_50.rds")
# saveRDS(best_parameters.null, "best_null.rds")
# saveRDS(best_parameters.L, "best_L.rds")
# saveRDS(best_parameters.R, "best_R.rds")
```

### Multiple counties, single change point - Nelder-Mead with all counties
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

eligible_counties <- readRDS("top1counties.rds")
df <- df |> filter(geoid_2010 %in% eligible_counties)
init_conds <- readRDS("init_conds_50.rds")$centers

best_parameters.null <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("ID", "nu", "nugget"))

errs <- c()
for (countyID in eligible_counties) {
      print(countyID)
      # pull their time series
      my_data <- df |> filter(geoid_2010 == countyID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      likelihood_wrapper.null <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
i <- 2      
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.null,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }

      best_parameters.null <- rbind(best_parameters.null, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      if (var(centered_rev.L) == 0){
            errs <- c(errs, countyID)
            next
            }
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.left,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      best_parameters.L <- rbind(best_parameters.L, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      if (var(centered_rev.R) == 0){
            errs <- c(errs, countyID)
            next
            }
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.right,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      best_parameters.R <- rbind(best_parameters.R, 
                                    as.data.frame(list(geoid = countyID,
                                                       nu = tst[which.max(tst$likelihood),]$nu,
                                                       nugget = tst[which.max(tst$likelihood),]$nugget)))

}

saveRDS(best_parameters.null, "best_null.rds")
saveRDS(best_parameters.L, "best_L.rds")
saveRDS(best_parameters.R, "best_R.rds")

```


```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

init_conds <- readRDS("init_conds_50.rds")$centers
countyID <- "21223" # "13061", "28143", "48431"

my_data <- df |> filter(geoid_2010 == countyID) |>
      mutate(TOT_REV = log1p(TOT_REV),
             TOT_EXP = log1p(TOT_EXP),
             TOT_ASSET = log1p(TOT_ASSET))
# min_year <- min(my_data$TAX_YEAR)
all_years <- my_data$TAX_YEAR - 1991 + 1 # all the years we have data for this county

candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - 1991 + 1
left_years <- all_years[1:which(all_years == candidate)-1]
right_years <- all_years[which(all_years == candidate):length(all_years)]

log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)

curr_all_years <- c(left_years, right_years)
centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))

i <- 1
mat_cov <- Matern(d = dist_mat, smoothness = init_conds[i,1], range = 1, phi = 1) + diag(init_conds[i,1], dim(dist_mat)[1])

# Compute sigma^2
sigma.squared <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(mat_cov)) %*% centered_rev)[1,1]

# Compute likelihood
likelihood <- dmvnorm(centered_rev,
                      mean = rep(0, length(centered_rev)),
                      sigma = sigma.squared * mat_cov,
                      log = TRUE)

centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
dist_mat.L <- dist_mat_all[left_years, left_years]
dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))

centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
dist_mat.R <- dist_mat_all[right_years, right_years]
dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))

mat_cov <- Matern(d = dist_mat.R, smoothness = init_conds[i,1], range = 1, phi = 1) + diag(init_conds[i,1], dim(dist_mat.R)[1])

# Compute sigma^2
sigma.squared <- (1/length(centered_rev.R)) * (centered_rev.R %*% chol2inv(chol(mat_cov)) %*% centered_rev.R)[1,1]

# Compute likelihood
likelihood <- dmvnorm(centered_rev.R,
                      mean = rep(0, length(centered_rev.R)),
                      sigma = sigma.squared * mat_cov,
                      log = TRUE)
```

## LRT
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

eligible_counties <- readRDS("top1counties.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top) |> filter(geoid_2010 %in% eligible_counties)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

best_all.null <- readRDS("best_null.rds")
best_all.L <- readRDS("best_L.rds")
best_all.R <- readRDS("best_R.rds")
rho = 1

reject_null <- list()
for (my_county in eligible_counties) {
      if (my_county == "21223"){ # this was one case where left was all 0s
            next
      }
      # pull their time series
      my_data <- df |> filter(geoid_2010 == my_county) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - min(df$TAX_YEAR) + 1
      
      # pull best parameters
      best.null <- best_all.null |> filter(geoid == my_county)
      best.L <- best_all.L |> filter(geoid == my_county)
      best.R <- best_all.R |> filter(geoid == my_county)
      

      # Determine the left and right time series corresponding to the current candidate change point
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Hypothesis Model ###
      curr_all_years <- c(left_years, right_years) # Redundant for single candidate
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years]) # center the log revenue values so they are mean 0
      
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      Sigma_0 <- Matern(d = dist_mat,
                        smoothness = best.null$nu,
                        range = rho,
                        phi = 1) + diag(best.null$nugget, dim(dist_mat)[1])
      
      sigma.sqrd_0 <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(Sigma_0)) %*% centered_rev)[1,1]
      Sigma_0 <- sigma.sqrd_0 * Sigma_0
      
      ### Alternative Hypothesis Model ###
      # center the log revenue values so they are mean 0
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      
      # compute and normalize distance matrices
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      # Compute Matern covariance matrices
      Sigma_L <- Matern(d = dist_mat.L,
                  smoothness = best.L$nu,
                  range = rho,
                  phi = 1) + diag(best.L$nugget, dim(dist_mat.L)[1])
      sigma.sqrd_L <- (1/length(left_years)) * (centered_rev.L %*% chol2inv(chol(Sigma_L)) %*% centered_rev.L)[1,1]
      Sigma_L <- sigma.sqrd_L * Sigma_L
      
      Sigma_R <- Matern(d = dist_mat.R,
                  smoothness = best.R$nu,
                  range = rho,
                  phi = 1) + diag(best.R$nugget, dim(dist_mat.R)[1])
      sigma.sqrd_R <- (1/length(right_years)) * (centered_rev.R %*% chol2inv(chol(Sigma_R)) %*% centered_rev.R)[1,1]
      Sigma_R <- sigma.sqrd_R * Sigma_R
      
      ### Likelihood Ratio Test ###
      L_0 <- dmvnorm(centered_rev, 
                     mean = rep(0, length(centered_rev)), 
                     sigma = Sigma_0,
                     log = TRUE)
      
      L_left <- dmvnorm(centered_rev.L, 
                        mean = rep(0, length(centered_rev.L)), 
                        sigma = Sigma_L,
                        log = TRUE)
      
      L_right <- dmvnorm(centered_rev.R, 
                         mean = rep(0, length(centered_rev.R)), 
                         sigma = Sigma_R,
                         log = TRUE)
      
      L_alt <- L_left + L_right

      
      # likelihood ratio test statistic
      lrt <- -2*L_0 + 2*L_alt
      doF <- 4 #  (mu1, mu2, Sigma1, Sigma2, nu1, nu2, nugget1, nugget2) vs (mu, Sigma, nu, nugget) 
      reject_null[my_county] <- (lrt > qchisq(p = 0.05, df = doF, lower.tail = FALSE))
}
# rm(list = setdiff(ls(), c("df", "eligible_counties", "gaps", "top_summary")))
sum(unlist(reject_null)) / length(reject_null) # 57.04% of the time we have enough evidence to reject the null
```


# Empirical Distribution County-Level

```{r}
library(readr)
library(dplyr)
library(tidyverse)

set.seed(2727)
eligible_counties <- readRDS("top1counties.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ"))) |>
      filter(geoid_2010 %in% eligible_counties) |>
      arrange(geoid_2010, TAX_YEAR)

df <- df |> filter(geoid_2010 != "21223") |> 
      group_by(geoid_2010) |>
      mutate(n = n(),
             row_num = row_number(),
             eligible = (row_num > 5 & row_num <= n-4)) |> ungroup()

candidates_all <- list()
for (i in 1:50) {
      candidates <- df |> group_by(geoid_2010) |>
            filter(eligible) |>
            slice_sample(n = 1) |>
            select(geoid_2010, TAX_YEAR)
      
      candidates_list <- candidates |> pull(TAX_YEAR) |> as.list()
      names(candidates_list) <- candidates |> pull(geoid_2010)
      
      candidates_all[[as.character(i)]] <- candidates_list
}
saveRDS(candidates_all, "random_candidates50.rds")

# indices <- df |> mutate(idx = row_number()) |>
#       group_by(geoid_2010) |>
#       filter(eligible) |>
#       slice_sample(n = 1) |>
#       pull(idx)
# 
# df$candidate <- FALSE
# df$candidate[indices] <- TRUE
# df |> filter(candidate) |> pull(eligible) |> sum()
# 
# df <- df |> select(-n, row_num, eligible)
# saveRDS(df, "empirical_df.rds")
```

## Hyperparameter Opt
### Multiple counties, single change point - Grid Search w/ 50 counties
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
source("helper.R")

candidates_all <- readRDS("random_candidates50.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ"))) |>
      filter(geoid_2010 %in% names(candidates_all$`1`)) |>
      arrange(geoid_2010, TAX_YEAR)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

set.seed(2727)
eligible_counties <- sample(unique(df$geoid_2010), size = 50)

combos <- expand.grid(nu = seq(from = 0.01, to = 3, by = 0.07475), 
                      nugget = seq(from = 0, to = 3, by = 0.1))

best_parameters.null <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ID", "nu", "nugget", "trial"))
best_parameters.L <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ID", "nu", "nugget", "trial"))
best_parameters.R <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("ID", "nu", "nugget", "trial"))

for (countyID in eligible_counties) {
      print(countyID)
      # pull their time series
      my_data <- df |> filter(geoid_2010 == countyID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      for (i in names(candidates_all)){
            candidate <- candidates_all[[i]][[countyID]] - min(df$TAX_YEAR) + 1
            left_years <- all_years[1:which(all_years == candidate)-1]
            right_years <- all_years[which(all_years == candidate):length(all_years)]
            
            ### Null Model ###
            curr_all_years <- c(left_years, right_years)
            centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
            dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
            dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
            
            # grid search (if doing Nelder-Mead, multiply get_likelihood_sigma by -1)
            likelihood_wrapper.null <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat,
                                                   rho = 1,
                                                   data = centered_rev))
            }
            
            likelihoods.null <- apply(combos, MARGIN = 1, likelihood_wrapper.null)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.null))
            best_parameters.null <- rbind(best_parameters.null, 
                                          as.data.frame(list(geoid = countyID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
            
            ### Alternative Model ###
            
            # Left
            centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
            dist_mat.L <- dist_mat_all[left_years, left_years]
            dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
            
            likelihood_wrapper.left <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat.L,
                                                   rho = 1,
                                                   data = centered_rev.L))
            }
            
            likelihoods.L <- apply(combos, MARGIN = 1, likelihood_wrapper.left)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.L))
            best_parameters.L <- rbind(best_parameters.L, 
                                          as.data.frame(list(geoid = countyID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
            
            # Right
            centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
            dist_mat.R <- dist_mat_all[right_years, right_years]
            dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
            
            likelihood_wrapper.right <- function(pars.vec){
                  return(get_likelihood_sigma(row = pars.vec,
                                                   dist_mat = dist_mat.R,
                                                   rho = 1,
                                                   data = centered_rev.R))
            }
            
            likelihoods.R <- apply(combos, MARGIN = 1, likelihood_wrapper.right)
            tst <- as.data.frame(list(nu = combos$nu, nugget = combos$nugget, likelihood = likelihoods.R))
            best_parameters.R <- rbind(best_parameters.R, 
                                          as.data.frame(list(geoid = countyID,
                                                             nu = tst[which.max(tst$likelihood),]$nu,
                                                             nugget = tst[which.max(tst$likelihood),]$nugget,
                                                             trial = i)))
      }
}

# TODO: should I do kmeans clustering per trial or just overall? for now just overall but ask team
all_best <- rbind(best_parameters.null, best_parameters.L, best_parameters.R) |> select(nu,nugget)
cl <- kmeans(all_best, 5, iter.max = 20, nstart = 3)
# saveRDS(cl, "empirical_init_conds_50.rds")
# cl$centers
```

### Multiple counties, single change point - Nelder-Mead with all counties
```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)
library(doParallel)
source("helper.R")

candidates_all <- readRDS("random_candidates50.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ"))) |>
      filter(geoid_2010 %in% names(candidates_all$`1`)) |>
      arrange(geoid_2010, TAX_YEAR)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

init_conds <- readRDS("empirical_init_conds_50.rds")$centers

best_parameters <- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "geoid", "nu", "nugget", "trial"))

errs <- c()
combos <- expand.grid(unique(df$geoid_2010), as.character(1:length(candidates_all))) |> arrange(Var1, Var2)

test_idx <- c(1:10, 51:60)
resBest <- foreach(i = test_idx, .combine = 'rbind', .packages = c("dplyr", "fields", "mvtnorm"), .verbose = TRUE, .errorhandling = "remove") %do% {
      countyID <- combos$Var1[i]
      trialID <- combos$Var2[i]
      print(paste("County ID:", countyID, "Trial Number:", trialID))
      # pull their time series
      my_data <- df |> filter(geoid_2010 == countyID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- candidates_all[[trialID]][[countyID]] - min(df$TAX_YEAR) + 1
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Model ###
      curr_all_years <- c(left_years, right_years)
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      
      likelihood_wrapper.null <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat,
                                             rho = 1,
                                             data = centered_rev))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
i <- 2      
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.null,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      res.null <- as.data.frame(list(type = "null",
                         geoid = countyID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID))
      
      ### Alternative Model ###
      
      # Left
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      if (var(centered_rev.L) == 0){
            cat(i, file = file.path("errors/", paste0(i, ".txt")))
            }
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      likelihood_wrapper.left <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.L,
                                             rho = 1,
                                             data = centered_rev.L))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.left,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      res.left <- as.data.frame(list(type = "left",
                         geoid = countyID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID))
      
      # Right
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      if (var(centered_rev.R) == 0){
            cat(i, file = file.path("errors/", paste0(i, ".txt")))
            }
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      likelihood_wrapper.right <- function(pars.vec){
            return(-1 * get_likelihood_sigma(row = pars.vec,
                                             dist_mat = dist_mat.R,
                                             rho = 1,
                                             data = centered_rev.R))
      }
      
      tst <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("nu.init", "nugg.init", "nu", "nugget", "likelihood", "trial"))
      for (i in 1:nrow(init_conds)){
            res <- constrOptim(theta = init_conds[i,], 
                               f = likelihood_wrapper.right,
                               grad = NULL,
                               ui = rbind(c(1,0),c(-1,0),c(0,1)), #rbind(c(1,0),c(-1,0),c(0,1))
                               ci = c(0.001,-3.5,0)) #c(0.001,-3.5, 0)
            tst[nrow(tst)+1,] <- c(c(init_conds[i,1], init_conds[i,2], res$par[1], res$par[2], -1*res$value, i))
      }
      rbind(res.null,
            res.left,
            as.data.frame(list(type = "right",
                         geoid = countyID,
                         nu = tst[which.max(tst$likelihood),]$nu,
                         nugget = tst[which.max(tst$likelihood),]$nugget,
                         trial = trialID)))

}

saveRDS(resBest, "empirical_best_test.rds")

```


```{r}
library(data.table)
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ")))

top <- quantile(df$TotPerCapADJ, probs = c(0.99)) # will look for county,year pairs where the county had per capita damage in the top 1 percentile
df <- df |> mutate(topDamage = TotPerCapADJ >= top)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

init_conds <- readRDS("init_conds_50.rds")$centers
countyID <- "21223" # "13061", "28143", "48431"

my_data <- df |> filter(geoid_2010 == countyID) |>
      mutate(TOT_REV = log1p(TOT_REV),
             TOT_EXP = log1p(TOT_EXP),
             TOT_ASSET = log1p(TOT_ASSET))
# min_year <- min(my_data$TAX_YEAR)
all_years <- my_data$TAX_YEAR - 1991 + 1 # all the years we have data for this county

candidate <- (my_data  |> filter(topDamage) |> select(TAX_YEAR) |> pull()) - 1991 + 1
left_years <- all_years[1:which(all_years == candidate)-1]
right_years <- all_years[which(all_years == candidate):length(all_years)]

log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)

curr_all_years <- c(left_years, right_years)
centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years])
dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))

i <- 1
mat_cov <- Matern(d = dist_mat, smoothness = init_conds[i,1], range = 1, phi = 1) + diag(init_conds[i,1], dim(dist_mat)[1])

# Compute sigma^2
sigma.squared <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(mat_cov)) %*% centered_rev)[1,1]

# Compute likelihood
likelihood <- dmvnorm(centered_rev,
                      mean = rep(0, length(centered_rev)),
                      sigma = sigma.squared * mat_cov,
                      log = TRUE)

centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
dist_mat.L <- dist_mat_all[left_years, left_years]
dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))

centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
dist_mat.R <- dist_mat_all[right_years, right_years]
dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))

mat_cov <- Matern(d = dist_mat.R, smoothness = init_conds[i,1], range = 1, phi = 1) + diag(init_conds[i,1], dim(dist_mat.R)[1])

# Compute sigma^2
sigma.squared <- (1/length(centered_rev.R)) * (centered_rev.R %*% chol2inv(chol(mat_cov)) %*% centered_rev.R)[1,1]

# Compute likelihood
likelihood <- dmvnorm(centered_rev.R,
                      mean = rep(0, length(centered_rev.R)),
                      sigma = sigma.squared * mat_cov,
                      log = TRUE)
```


## LRT
```{r}
library(readr)
library(dplyr)
library(tidyverse)

glued <- rbind(readRDS( "best_empirical_1stHalf.rds"), 
               readRDS( "best_empirical_2ndHalf.rds"))

best_all <- glued |> 
      mutate(ID = paste0(geoid,".",trial)) |> 
      select(-geoid,-trial)
best_all <- split(best_all, best_all$ID)
best_all <- lapply(best_all, function(x) {split(x,x$type)})

saveRDS(best_all, "best_empirical_all.rds")
```


```{r}
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

candidates_all <- readRDS("random_candidates50.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ"))) |>
      filter(geoid_2010 %in% names(candidates_all$`1`)) |>
      arrange(geoid_2010, TAX_YEAR)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

best_all <- readRDS("best_empirical_all.rds")

rho = 1

reject_null_all <- setNames(vector("list", length(candidates_all)), names(candidates_all))
for (trialID in 1:length(candidates_all)) {
      reject_null <- list()
      for (countyID in unique(df$geoid_2010)){
            ID <- paste0(countyID, ".", trialID)
            print(ID)
            # pull their time series
            my_data <- df |> filter(geoid_2010 == countyID) |>
                  mutate(TOT_REV = log1p(TOT_REV),
                         TOT_EXP = log1p(TOT_EXP),
                         TOT_ASSET = log1p(TOT_ASSET))
            all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
            log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
            
            candidate <- candidates_all[[trialID]][[countyID]] - min(df$TAX_YEAR) + 1
            
            # pull best parameters
            best.curr <- best_all[[ID]] # best.curr$left$nu
      
            # Determine the left and right time series corresponding to the current candidate change point
            left_years <- all_years[1:which(all_years == candidate)-1]
            right_years <- all_years[which(all_years == candidate):length(all_years)]
            
            ### Null Hypothesis Model ###
            curr_all_years <- c(left_years, right_years) # Redundant for single candidate
            centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years]) # center the log revenue values so they are mean 0
            
            dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
            dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
            Sigma_0 <- Matern(d = dist_mat,
                              smoothness = best.curr$null$nu,
                              range = rho,
                              phi = 1) + diag(best.curr$null$nugget, dim(dist_mat)[1])
            
            sigma.sqrd_0 <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(Sigma_0)) %*% centered_rev)[1,1]
            Sigma_0 <- sigma.sqrd_0 * Sigma_0
            
            ### Alternative Hypothesis Model ###
            # center the log revenue values so they are mean 0
            centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
            centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
            
            # compute and normalize distance matrices
            dist_mat.L <- dist_mat_all[left_years, left_years]
            dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
            
            dist_mat.R <- dist_mat_all[right_years, right_years]
            dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
            
            # Compute Matern covariance matrices
            Sigma_L <- Matern(d = dist_mat.L,
                        smoothness = best.curr$left$nu,
                        range = rho,
                        phi = 1) + diag(best.curr$left$nugget, dim(dist_mat.L)[1])
            sigma.sqrd_L <- (1/length(left_years)) * (centered_rev.L %*% chol2inv(chol(Sigma_L)) %*% centered_rev.L)[1,1]
            Sigma_L <- sigma.sqrd_L * Sigma_L
            
            Sigma_R <- Matern(d = dist_mat.R,
                        smoothness = best.curr$right$nu,
                        range = rho,
                        phi = 1) + diag(best.curr$right$nugget, dim(dist_mat.R)[1])
            sigma.sqrd_R <- (1/length(right_years)) * (centered_rev.R %*% chol2inv(chol(Sigma_R)) %*% centered_rev.R)[1,1]
            Sigma_R <- sigma.sqrd_R * Sigma_R
            
            ### Likelihood Ratio Test ###
            L_0 <- dmvnorm(centered_rev, 
                           mean = rep(0, length(centered_rev)), 
                           sigma = Sigma_0,
                           log = TRUE)
            
            L_left <- dmvnorm(centered_rev.L, 
                              mean = rep(0, length(centered_rev.L)), 
                              sigma = Sigma_L,
                              log = TRUE)
            
            L_right <- dmvnorm(centered_rev.R, 
                               mean = rep(0, length(centered_rev.R)), 
                               sigma = Sigma_R,
                               log = TRUE)
            
            L_alt <- L_left + L_right
      
            
            # likelihood ratio test statistic
            lrt <- -2*L_0 + 2*L_alt
            doF <- 4 #  (mu1, mu2, Sigma1, Sigma2, nu1, nu2, nugget1, nugget2) vs (mu, Sigma, nu, nugget)
            reject_null[[countyID]] <- (lrt > qchisq(p = 0.05, df = doF, lower.tail = FALSE))

      }
      reject_null_all[[as.character(trialID)]] <- reject_null
}

# res <- unlist(lapply(reject_null_all, function(x){sum(unlist(x)) / length(x)}))
# hist(res)
# sd(res)
```

bootstrapping
```{r}
library(doParallel)

res <- unlist(unname(reject_null_all))
res <- split(res, names(res))
res <- lapply (res, unname)

empirical <- foreach(i = 1:1000, .inorder = FALSE) %do% {
      tst <- lapply(res, function(x){sample(x, 1)})
      sum(unlist(tst)) / length(tst)
}

hist(unlist(empirical))
quantile(unlist(empirical), probs = c(0.025, 0.5, 0.975)) # middle 95% of data is between 50.54% and 58.49%
mean(unlist(empirical))
```


```{r}
library(readr)
library(dplyr)
library(tidyverse)
library(fields)
library(mvtnorm)

candidates_all <- readRDS("random_candidates50.rds")
df <- readRDS("county_long.rds") |> 
      filter(TAX_YEAR < 2020) |> 
      select(all_of(c("TAX_YEAR", "geoid_2010", "TOT_REV", "TOT_EXP", "TOT_ASSET", "county_full", "state_name", "REGION", "DIVISION", "TotDmgADJ", "n_disasters", "TotPerCapADJ", "FEMA_threshADJ"))) |>
      filter(geoid_2010 %in% names(candidates_all$`1`)) |>
      arrange(geoid_2010, TAX_YEAR)

# distance matrix of years
dist_mat_all <- as.matrix(dist(sort(unique(df$TAX_YEAR) - min(df$TAX_YEAR) + 1), diag=TRUE, upper=TRUE))

best_all <- readRDS("empirical_best_test.rds") 
combos <- expand.grid(unique(best_all$geoid), unique(best_all$trial)) |> arrange(Var1, Var2)
reject_null <- setNames(vector("list", length(unique(best_all$trial))), unique(best_all$trial))
reject_null <- lapply(reject_null, function(x){
      setNames(vector("list", length(unique(best_all$geoid))), unique(best_all$geoid))
      })

best_all <- best_all |> 
      mutate(ID = paste0(geoid,".",trial)) |> 
      select(-geoid,-trial)
best_all <- split(best_all, best_all$ID)
best_all <- lapply(best_all, function(x) {split(x,x$type)})

rho = 1

# combos <- expand.grid(unique(df$geoid_2010), as.character(1:length(candidates_all))) |> arrange(Var1, Var2)
for (i in nrow(combos)) {
      countyID <- combos$Var1[i]
      trialID <- combos$Var2[i]
      ID <- paste0(countyID, ".", trialID)
      print(ID)
      # pull their time series
      my_data <- df |> filter(geoid_2010 == countyID) |>
            mutate(TOT_REV = log1p(TOT_REV),
                   TOT_EXP = log1p(TOT_EXP),
                   TOT_ASSET = log1p(TOT_ASSET))
      all_years <- my_data$TAX_YEAR - min(df$TAX_YEAR) + 1 # all the years we have data for this county
      log_rev <- my_data |> merge(data.frame(TAX_YEAR = 1991:2019), all = TRUE) |> arrange(TAX_YEAR) |> pull(TOT_REV)
      
      candidate <- candidates_all[[trialID]][[countyID]] - min(df$TAX_YEAR) + 1
      
      # pull best parameters
      best.curr <- best_all[[ID]] # best.curr$left$nu

      # Determine the left and right time series corresponding to the current candidate change point
      left_years <- all_years[1:which(all_years == candidate)-1]
      right_years <- all_years[which(all_years == candidate):length(all_years)]
      
      ### Null Hypothesis Model ###
      curr_all_years <- c(left_years, right_years) # Redundant for single candidate
      centered_rev <- log_rev[curr_all_years] - mean(log_rev[curr_all_years]) # center the log revenue values so they are mean 0
      
      dist_mat <- dist_mat_all[curr_all_years, curr_all_years]
      dist_mat <- (dist_mat - min(dist_mat)) / (max(dist_mat) - min(dist_mat))
      Sigma_0 <- Matern(d = dist_mat,
                        smoothness = best.curr$null$nu,
                        range = rho,
                        phi = 1) + diag(best.curr$null$nugget, dim(dist_mat)[1])
      
      sigma.sqrd_0 <- (1/length(centered_rev)) * (centered_rev %*% chol2inv(chol(Sigma_0)) %*% centered_rev)[1,1]
      Sigma_0 <- sigma.sqrd_0 * Sigma_0
      
      ### Alternative Hypothesis Model ###
      # center the log revenue values so they are mean 0
      centered_rev.L <- log_rev[left_years] - mean(log_rev[left_years])
      centered_rev.R <- log_rev[right_years] - mean(log_rev[right_years])
      
      # compute and normalize distance matrices
      dist_mat.L <- dist_mat_all[left_years, left_years]
      dist_mat.L <- (dist_mat.L - min(dist_mat.L)) / (max(dist_mat.L) - min(dist_mat.L))
      
      dist_mat.R <- dist_mat_all[right_years, right_years]
      dist_mat.R <- (dist_mat.R - min(dist_mat.R)) / (max(dist_mat.R) - min(dist_mat.R))
      
      # Compute Matern covariance matrices
      Sigma_L <- Matern(d = dist_mat.L,
                  smoothness = best.curr$left$nu,
                  range = rho,
                  phi = 1) + diag(best.curr$left$nugget, dim(dist_mat.L)[1])
      sigma.sqrd_L <- (1/length(left_years)) * (centered_rev.L %*% chol2inv(chol(Sigma_L)) %*% centered_rev.L)[1,1]
      Sigma_L <- sigma.sqrd_L * Sigma_L
      
      Sigma_R <- Matern(d = dist_mat.R,
                  smoothness = best.curr$right$nu,
                  range = rho,
                  phi = 1) + diag(best.curr$right$nugget, dim(dist_mat.R)[1])
      sigma.sqrd_R <- (1/length(right_years)) * (centered_rev.R %*% chol2inv(chol(Sigma_R)) %*% centered_rev.R)[1,1]
      Sigma_R <- sigma.sqrd_R * Sigma_R
      
      ### Likelihood Ratio Test ###
      L_0 <- dmvnorm(centered_rev, 
                     mean = rep(0, length(centered_rev)), 
                     sigma = Sigma_0,
                     log = TRUE)
      
      L_left <- dmvnorm(centered_rev.L, 
                        mean = rep(0, length(centered_rev.L)), 
                        sigma = Sigma_L,
                        log = TRUE)
      
      L_right <- dmvnorm(centered_rev.R, 
                         mean = rep(0, length(centered_rev.R)), 
                         sigma = Sigma_R,
                         log = TRUE)
      
      L_alt <- L_left + L_right

      
      # likelihood ratio test statistic
      lrt <- -2*L_0 + 2*L_alt
      doF <- 4 #  (mu1, mu2, Sigma1, Sigma2, nu1, nu2, nugget1, nugget2) vs (mu, Sigma, nu, nugget)
      print(reject_null[[trialID]][[countyID]])
      reject_null[[trialID]][[countyID]] <- (lrt > qchisq(p = 0.05, df = doF, lower.tail = FALSE))
      print(reject_null[[trialID]][[countyID]])
}


# sum(unlist(reject_null)) / length(reject_null) # 52.71%
```


